---
title: "ERA"
author: "Will Kolbrener"
date: "2023-12-02"
output: pdf_document
---

Data Cleaning
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(caret)


teams <- read.csv("Teams.csv")

teams <- teams %>% filter(yearID >= 1980)

Team_Full_Data <- teams

teams$teamID <- NULL

teams$ER <- NULL

teams$HRA <- NULL

teams$RA <- NULL

teams$IPouts <- NULL

teams$park <- NULL

teams$HA <- NULL

teams$DivWin <- NULL

teams$WCWin <- NULL

teams$LgWin <- NULL

teams$WSWin <- NULL

teams$teamID<- NULL

teams$franchID <- NULL

teams$names <- NULL

teams$HBP <- NULL

teams$SF <- NULL

teams$name <- NULL

teams$teamIDBR <- NULL

teams$teamIDlahman45 <- NULL

teams$teamIDretro <- NULL

teams$Rank <- NULL

teams$G <- NULL

teams$L <- NULL

teams$AB <- NULL

teams$H <- NULL

teams$HR <- NULL

teams$BPF <- NULL 

teams$G <- NULL 


#NA Check
length(which(is.na(teams==TRUE)))
teams[which(is.na(teams==TRUE)),]


Predic_ERA <- teams



```


Model Construction
```{r}

nTotal<- nrow(Predic_ERA)

nTrain<-814

nTest<-204


set.seed(542)
train<-sample(nTotal, nTrain)

model<-lm(ERA~., data=Predic_ERA[train,]) #builds linear model just off training data


summary(model)


vif(model)



## testing
pred<-predict(model, newdata <-Predic_ERA[-train,])# makes prediction (applies model) on just 





diff<-Predic_ERA$ERA[-train]-pred

error<-sqrt(sum(diff**2)/nTest)

cat(sprintf('nTrain=%d, nTest=%d, error=%f\n', nTrain, nTest, error))

order(diff,decreasing=TRUE)

max(diff)

newdata$diff<- diff

newdata$predictedERA<- pred


newdata[200,]


extracted_rows <- teams[rownames(newdata), ]


```


Cross Validation
```{r}
fit <- train(ERA ~ ., data=Predic_ERA, method='lm', trControl = trainControl(method='cv', number=10))
fit
```


Predicted ERA vs Observed ERA
```{r}

newdata$Type <- rep("Observed", nrow(newdata))
newdata$Type[which(!is.na(newdata$predictedERA))] <- "Predicted"

ggplot(newdata, aes(x=yearID)) +
  geom_smooth(aes(y=ERA, color="Observed"), se=FALSE) +
  geom_smooth(aes(y=predictedERA, color="Predicted"), se=FALSE) +
  labs(x="Year", y="ERA", title="Predicted ERA vs Observed ERA") +
  scale_color_manual(values=c("Observed"="red", "Predicted"="black")) +
  theme(legend.title=element_blank()) # Optional: remove the legend title if desired


```



Biggest Misses
```{r}

extracted_rows <- Team_Full_Data[rownames(newdata), ]
order(abs(diff),decreasing=TRUE)

#83, 104, 142, 124, 107

#1994 Minnesota Twin
extracted_rows[83,]
newdata[83,]

#1998 KC Royals
extracted_rows[104,]
newdata[104,]

# 2005 Arizona Diamondbacks
extracted_rows[142,]
newdata[142,]


# 2001 New York Mets
extracted_rows[124,]
newdata[124,]


# 1998 Philadelphia Phillies
extracted_rows[107,]
newdata[107,]


View(Team_Full_Data %>% filter(teamID == "NYN"))


ggplot(Team_Full_Data,aes(x=R , y=ERA)) + geom_smooth()

teamsw <- Team_Full_Data %>% filter(teamID =="PHI" & yearID==1998)
median(teamsw$W)
      
```
1. 1994 Minnesota Twins: Observed ERA: 5.68 Predicted ERA: 4.858362
2. 1998 Kansas City Royals: Observed ERA: 5.16 Predicted ERA: 4.574792
3. 2005 Arizona Diamondbacks: Observed ERA: 4.84 Predicted ERA:4.273108
4. 2001 New York Mets: Observed ERA: 4.07 Predicted ERA: 3.505965
5. 1998 Philadelphia Phillies: Observed ERA: 4.64 Predicted ERA: 4.175759



Potential Reason for large miss of the 1994 MN Twins

- Wins, Runs and Saves are all significant
- 1994 MLB regular season ended 6 weeks early due to a players strike, this resulted in the Twins playing 49 less games than they have played in past years resulting in incomplete season statistics 
-1993 Twins won 71 games where the 1994 Twins won 53
-1993 Twins Recorded 44 saves and the 1994 Twins recorded 29 saves 
-1993 Twins Scored 693 runs and the 1994 Twins scored 594


Potential Reason for large miss of the 1998 KC Royals
- Wins, Runs and Saves are all significant
- Royals Record only 65 wins, the MLB median number of wins was 79.5. Teams with more wins tend to have lower ERAs 


Potential Reason for large miss of the 2005 Diamondback
- Ended the season 25th in Era (Had the 25th highest ERA in MLB)
- Diamondbacks pitcher Brandon Webb won the Cy Young Award in 2005.
-The average ERA of past Cy Young Winners is 2.48, Brandon Webb won the award with an ERA of 3.10
- Diamondbacks hired a new manager at the start of the seaosn  


Potential Reason for large miss of 2001 NY Mets
- Had a large number of wins (82) but pitching staff preformed poorly
- In 2000 the Mets scored 807 runs but in 2001 the Mets scored 642 runs, this large decrease in runs scored could have impacted this prediction
- Mets had many home games rescheduled due to the September 11th Attack 


Potential Reason for large miss of the 1998 Philadelphia Phillies
 - First year in new ball park, which is extremely hitters friendly
 - Team had many offensive stars such as Scott Rolen 
 - Star pitchers such as Curt Schillings had a poor preformences in the new park 


