---
title: "DataMiningProject_EDA"
author: "Ian Leonard"
date: "2023-11-28"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(caret)
```

```{r}
master <- read.csv("Master.csv")
batting <- read.csv("Batting.csv")
salaries <- read.csv("Salaries.csv")
teams <- read.csv("Teams.csv")
pitching <- read.csv("Pitching.csv")
allstar <- read.csv("AllstarFull.csv")
```

Summaries of Interesting Cols

Batting:
 - 5149 NA's (0.05081317) in AB
 - Mean: 150 ABs
 
Salary:
 - No entries for salary before 1985

```{r}
summary(batting$AB)
# count of NA's in batting
sum(is.na(batting$AB))
nrow(batting)
```
All Stars
```{r}
summary(allstar$GP)

```

Important Cols: 
 - PlayerID, YearID, 
 
 
1. Predict All Star Appearances based on Batting Stats
```{r}
# merge allstar and batting

# take 
allstar_batting <- batting %>% 
  left_join(allstar, by = c("playerID", "yearID"))

summary(allstar_batting)
# remove NA's

allstar_batting <- na.omit(allstar_batting)
```

2. Predict Salary based on Batting Stats
```{r}
# create table with playerID, yearID, salary, batting stats
batting_salary <- batting %>% 
  left_join(salaries, by = c("playerID", "yearID"))

#plot average salary by year
#ggplot(batting_salary, aes(yearID, salary)) + geom_point() + geom_smooth(method = "lm")

year_salary <- batting_salary %>% 
  group_by(yearID) %>% 
  summarise(mean_salary = mean(salary, na.rm = TRUE)) %>% 
  filter(!yearID < 1985)

ggplot(year_salary, aes(yearID, mean_salary)) + geom_point() + geom_smooth(method = "lm")
```
Model for predicting salary based on batting stats
FOR YEAR 1985-1990

Notes: 
 - 1985-1990 has 1918/6310 = 0.304 NA's in salary
```{r}

batting_salary_80s <- batting_salary %>% 
  filter(yearID >= 1985 & yearID <= 1990)
sum(is.na(batting_salary_80s$salary))
nrow(batting_salary_80s)

```

1990-2000
NOTES:
  - 2873/13373 = 0.2148359 NA's in salary
```{r}
batting_salary_90s <- batting_salary %>% 
  filter(yearID >= 1990 & yearID <= 2000)
sum(is.na(batting_salary_90s$salary))
nrow(batting_salary_90s)



```

2000-2010
Notes:
  - 5116/14959 = 0.3420015 NA's in salary
```{r}
batting_salary_00s <- batting_salary %>% 
  filter(yearID >= 2000 & yearID <= 2010)
sum(is.na(batting_salary_00s$salary))
nrow(batting_salary_00s)

```

2010-2015
Notes:
  - 3136/8485 = 0.3695934 NA's in salary

```{r}
batting_salary_10s <- batting_salary %>% 
  filter(yearID >= 2010 & yearID <= 2015)
sum(is.na(batting_salary_10s$salary))
nrow(batting_salary_10s)

```

```{r}
batting_salary_80s <- na.omit(batting_salary_80s)
# create model
batting_salary_80s_model <- lm(salary ~ G + AB + R + H + X2B + X3B + HR + RBI + SB + CS + BB + SO + IBB + HBP + SH + SF + GIDP, data = batting_salary_80s)
summary(batting_salary_80s_model)
```

```{r}
```

3. Predict All Star Appearances based on Batting, Pitching, and Fielding Stats


4. Predict Batting Average for 2015 based on stats from 2014
Batting Average = H/AB
```{r}
# create table with playerID, yearID, salary, batting stats from 2014
batting_2014 <- batting_salary %>% 
  filter(yearID == 2014)
# insert Batting Average
batting_2014$BA <- batting_2014$H/batting_2014$AB

batting_2015 <- batting_salary %>% 
  filter(yearID == 2015)
# insert Batting Average
batting_2015$BA <- batting_2015$H/batting_2015$AB

# create model
batting_2014_model <- lm(BA ~ H+ G + AB + R + X2B + X3B + HR + RBI + SB + CS + BB + SO + IBB + HBP + SH + SF + GIDP, data = batting_2014)
summary(batting_2014_model)
```

```{r}
# Install and load the rpart package
library(rpart)

# Create a decision tree model
batting_2014_tree <- rpart(BA ~ H + G + AB + R + X2B + X3B + HR + RBI + SB + CS + BB + SO + IBB + HBP + SH + SF + GIDP, data = batting_2014)

# Print the decision tree
print(batting_2014_tree)

# Plot the decision tree
plot(batting_2014_tree)
text(batting_2014_tree, cex = 0.8)

# Make predictions for 2015 using the model
batting_2015_predictions <- predict(batting_2014_tree, newdata = batting_2015)  # Assuming you have a batting_2015 dataset

# Add the predictions to your batting_2015 dataset
batting_2015$Predicted_BA <- batting_2015_predictions

# View the predictions
batting_2015$Predicted_BA

# View the actual batting averages
batting_2015$BA

# Calculate the mean squared error
mean((batting_2015$BA - batting_2015$Predicted_BA)^2)
```



